---
subtitle: 'Developing Data Products - Course Project'
title: "Shiny Application and Reproducible Pitch"
author: "Antonio Vitor"
date: "March 16th 2022"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(shiny)
library(dplyr)
library(plotly)

data(mtcars)

mtcars <- mtcars %>%
     mutate(am = factor(am,
                        labels = c("Automatic", "Manual")))
```

## Introduction

This project is the final peer assessment for the [Developing Data Products](https://www.coursera.org/learn/data-products?specialization=jhu-data-science) course from the Johns Hopkins Data Science Specialization.

The Shiny Application is available at: <br>
https://antoniovitor.shinyapps.io/jhu-ddp-shinyapplication/

The source files for this projects are available at: <br>
https://github.com/antoniovitorvb/JHU_Data_Science/tree/main/9%20Developing%20Data%20Products/week4/CourseProject

## Application Overview

In this app it is possible to manipulate the MPG ~ Weight scatter plot of the `mtcars` dataset by selecting the range of MPG.

It can also show the regression line for the data points available separated by the transmission type of the cars.

## How the app works

Side Panel:  
- Select the range of MPG  
- Select if you want the regression line to be shown  

Main Panel:  
- See the scatter plot with the corresponding range of MPG  
- See (or not) the regression line for each type of transmission  

## Demonstration

Here is an example of the plot without the user input values

```{r, message=FALSE}
gg <- ggplot(mtcars,
             aes(x = wt,
                 y = mpg,
                 color = am)) +
     geom_point() +
     geom_smooth(method = "lm") +
     labs(x = "Weight (1000 lbs)",
          y = "MPG",
          color = "Transmission Type")

ggplotly(gg)
```

