---
title: "Regression Models Course Project"
author: "Antonio Vitor Villas Boas"
date: "07/02/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Executive Summary

In this article it will be presented an analysis comparing automatic and manual transmission impacts on Miles Per Gallon (MPG) and quantify the MPG difference difference between the two transmissions.

The data set to be studied is the `mtcars`. The data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models).


```{r warning=FALSE, message=FALSE, cache=TRUE}
library(ggplot2)
library(dplyr)
library(regclass)

data(mtcars)

mtcars$am <- factor(mtcars$am, labels = c("Automatic", "Manual"))
```

Now let's make some exploratory analysis:

```{r}
ggplot(mtcars,
       aes(x = am,
           y = mpg,
           fill = am)) +
     geom_boxplot() +
     labs(x = "", fill = "")

fit <- lm(mpg ~ am, mtcars)
summary(fit)
```

Considering only the transmission as a predictor we can conclude that manual transmission cars have a better MPG performance over automatic transmission ones.

Although there are other variables that may impact this conclusion such as weight, number of cylinders, gross horsepower etc. Let's find out then which ones can truly impact on MPG.

```{r, cache=TRUE}
full_fit <- lm(mpg ~ ., mtcars)

summary(full_fit)
```

It is important to have in mind that excluding correlated variables can lead to biased conclusions, including uncorrelated predictors can inflate the variance of the model. To measure this inflation, the `VIF` function calculates the Variation Inflation Factors of all predictors in a regressions model.

First let's use a different fit to the model as an comparison to `full_fit`. One with and extra predictors: `cyl`.

```{r}
fit2 <- lm(mpg ~ am + cyl, mtcars)
VIF(fit2)

VIF(full_fit)
```

comparing both am Variation Inflation Factors it shows that `am` is much more inflated when using all predictors in the model.

The function `step` in useful to select which predictors are better correlated to the model with the lowest Akaike Information Criterion (AIC).

```{r, cache=TRUE}
final_fit <- step(full_fit, direction = "backward", trace = F)
final_fit

VIF(final_fit)
```

This shows that the lowest IAC = 61.31 and the VIF = `r VIF(final_fit)[3]` using three predictors: `wt`, `qsec` and `am`.

With these coefficients, we can conclude that Manual transmission cars get 2.936 more MPG on average than automatic ones.

the confidence interval for this statment is:
```{r}
confint(final_fit)['amManual',]
```


\newpage
# Appendix

mpg ~ cyl with number of cylinders considered.
```{r}
mtcars$cyl <- factor(mtcars$cyl)

ggplot(mtcars,
       aes(x = cyl,
           y = mpg,
           fill = am)) +
     geom_boxplot() +
     labs(fill = "Transmission",
          x = "Number of Cylinders")
```

Null Hypothesis test
```{r}
t.test(mpg~am, mtcars)
```

